---
title: "linear regression"
author: "chad allison"
date: '2023-01-23'
output: github_document
---

### setup

```{r message = F, warning = F}
library(tidyverse)
library(class)
library(VGAM)
library(gridExtra)
theme_set(theme_minimal())
options(scipen = 999)
```

### data import

```{r message = F, warning = F}
df = read_csv("loan_data.csv", col_types = cols())
head(df)
```

### visualizing `loan_status` observations

```{r}
df |>
  count(loan_status) |>
  ggplot(aes(loan_status, n)) +
  geom_col(aes(fill = loan_status), alpha = 0.75) +
  geom_text(aes(label = n), vjust = -0.5) +
  scale_fill_manual(values = c("#DC7777", "#76A877")) +
  labs(x = "loan status", y = "count",
       title = "counts of loan statuses") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")
```

### visualizing `Principal`

```{r}
df |>
  ggplot(aes(Principal)) +
  geom_histogram(aes(fill = loan_status, y = 0.5 * ..density..),
                 binwidth = 120, alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("indianred3", "springgreen4"))
  labs(x = "principal", y = "density", fill = "loan status",
       title = "histogram of principal by loan status") +
  theme(plot.title = element_text(hjust = 0.5))
```

### visualizing `terms`

```{r warning = F}
df |>
  ggplot(aes(terms)) +
  geom_histogram(aes(fill = loan_status, y = 1 * ..density..),
                 binwidth = 10, alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("indianred3", "springgreen4")) +
  scale_x_continuous(limits = c(0, 40)) +
  labs(x = "terms", y = "density", fill = "loan status",
       title = "plot of loan status by terms") +
  theme(plot.title = element_text(hjust = 0.5))
```

### visualizing `age`

```{r}
df |>
  ggplot(aes(age)) +
  geom_histogram(aes(fill = loan_status, y = 1 * ..density..),
                 binwidth = 1, alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("indianred3", "springgreen4")) +
  labs(y = "density", fill = "loan status",
       title = "relationship between age and loan status") +
  theme(plot.title = element_text(hjust = 0.5))
```

### visualizing multiple variables

```{r warning = F}
hist_top = df |>
  ggplot(aes(Principal)) +
  geom_histogram(aes(fill = loan_status, y = 1 * ..density..),
                  binwidth = 100, alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c("indianred3", "springgreen4")) +
  scale_x_continuous(limits = c(200, 1100)) +
  labs(x = "principal", y = "density") +
  theme(legend.position = "none")

empty = ggplot() +
  geom_point(aes(1, 1), col = "white") +
  theme(axis.ticks = element_blank(), 
        panel.background = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

scatter = df |>
  ggplot(aes(Principal, age)) +
  geom_point(aes(col = loan_status, fill = loan_status)) +
  scale_color_manual(values = c("indianred3", "springgreen4")) +
  theme(legend.position = "top") +
  labs(x = "principal", y = "age")

hist_right = df |>
  ggplot(aes(age)) +
  geom_histogram(aes(fill = loan_status, y = 0.5 * ..density..),
                 binwidth = 1, alpha = 0.5, position = "identity") +
  scale_x_continuous(limits = c(20, 45)) +
  scale_fill_manual(values = c("indianred3", "springgreen4")) +
  coord_flip() +
  theme(legend.position = "none") +
  labs(x = "age", y = "density")


grid.arrange(hist_top, empty, scatter, hist_right,
             ncol = 2, nrow = 2, widths = c(4, 1), heights = c(1, 4))
```

### visualizing the day of week

```{r warning = F}
df |>
  ggplot(aes(dayofweek)) +
  geom_histogram(aes(fill = loan_status, y = 1 * ..density..),
                 binwidth = 1, alpha = 0.5, position = "identity") +
  scale_x_continuous(limits = c(0, 7)) +
  scale_fill_manual(values = c("indianred3", "springgreen4")) +
  labs(x = "day of week", y = "density", fill = "loan status",
       title = "loan status by day of week") +
  theme(plot.title = element_text(hjust = 0.5))
```

### converting days of week to categorical variable

```{r}
namevector = c("Weekend")
df[, namevector] = 0
df$Weekend[df$dayofweek > 3] = 1
head(select(df, dayofweek, Weekend))
```

### encoding `Gender`

```{r}
namevector = c("Gender01")
df[, namevector] = 0
df$Gender01[df$Gender == "male"] = 1
head(select(df, Gender, Gender01))
```

### table of gender and loan status

```{r}
table(df$Gender01, df$loan_status)
```

### visualizing relationship between gender and loan status

```{r warning = F}
df |>
  ggplot(aes(Gender01)) +
  geom_histogram(aes(fill = loan_status, y = 1 * ..density..),
                 binwidth = 1, alpha = 0.5, position = "identity") +
  scale_x_continuous(limits = c(0, 2)) +
  scale_fill_manual(values = c("indianred3", "springgreen4")) +
  labs(x = "gender", y = "density", fill = "loan status",
       title = "loan status by gender") +
  theme(plot.title = element_text(hjust = 0.5))
```

### one hot encoding education

```{r}
df = df |>
  mutate(educationBechalor = ifelse(education == "Bechalor", 1, 0),
         educationcollege = ifelse(education == "college", 1, 0),
         `educationHigh School or Below` = ifelse(education == "High School or Below", 1, 0),
         `educationMaster or Above` = ifelse(education == "Master or Above", 1, 0))

df |>
  select(14:17) |>
  head()
```

### getting data in data frame

```{r}
Columns = c("Principal", "terms", "age", "educationBechalor", "educationcollege",
            "educationHigh School or Below", "educationMaster or Above", "Weekend", "Gender01")

Data = df[Columns]
head(Data)
```

### adding labels

```{r}
NewColumn = c("Class")
Data[, NewColumn] = 0
Data$Class[df$loan_status == "PAIDOFF"] = 1
sample_n(select(Data, Class), 10)
```

### normalize data

```{r}
Data[Columns] = scale(Data[Columns])
head(Data[Columns])
```

### train test split

```{r}
set.seed(123)
testindex = sample.int(nrow(Data))[1:floor(0.1 * nrow(Data))]
TestData = Data[testindex, ]
TrainData = Data[-testindex, ]
paste0(nrow(TrainData), " observations in training data; ",
       nrow(TestData), " observations in testing data")
```

### building model

```{r}
model = glm(Class ~ ., family = binomial(link = "logit"), data = TrainData, control = list(maxit = 50))
summary(model)
```

### making predictions

```{r warning = F}
fitted.results = predict(model, newdata = TestData, type = "response")
yhat = ifelse(fitted.results > 0.5, 1, 0)
yhat[1:5]
```

### getting labels

```{r}
y = TestData[, c("Class")]
y[1:4, ]
```

### getting accuracy

```{r}
mean(yhat == y)
```






























































